---
title: "An Introduction to spARI"
author: "Yinqiao Yan"
output: 
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
date: "2025-07-15"
vignette: >
  %\VignetteIndexEntry{An Introduction to spARI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The R package **spARI** is designed to compute two novel clustering evaluation metrics—the **spatially aware Rand index (spRI)** and its adjusted version (**spARI**)—for assessing spatial transcriptomics clustering. Unlike the traditional Rand index (RI) and adjusted Rand index (ARI), spRI and spARI incorporate spatial distance information into clustering evaluation. When comparing two partitions, spRI assigns a weight to each disagreement pair—two objects in the same cluster of one partition but in different clusters of the other—based on their spatial distance, allowing for a more refined distinction between disagreement pairs. The spRI value ranges between zero and one, and the spARI value is less than or equal to one with an expected value of zero.  

To implement this package, users are required to input two partitions and either spatial coordinates or a precomputed distance matrix. The main function "spARI" then automatically computes spRI and spARI in a fast and user-friendly manner. If spatial coordinates are provided, a built-in normalization step is applied to remove the effect of spatial location unit. If a distance matrix is provided, for large-scale spatial data (e.g., 100,000 objects), users can input a sparse distance matrix of class "dgCMatrix" or "dgTMatrix", for instance, by retaining only distances to the $k$ nearest neighbors for each object. The `spARI` package is compatible with Windows, Linux, and macOS, and can be easily installed on all three platforms.

## Installation

Users can install the `spARI` package from GitHub.

```{r install, eval=FALSE}
devtools::install_github("yinqiaoyan/spARI")
```

Other required packages also need to be installed in advance.

```{r install2, eval=FALSE}
install.packages("FNN")  # find k-nearest neighbors
```

## Examples of using the package

### Compute spRI and spARI using the spatial coordinates

The following code shows an example that runs the main function "spARI" in our package.

Import the `spARI` package and load the example data stored in this package.

```{r}
library(spARI)
data("spARI_example_data")
true_labels = spARI_example_data$true_labels
c1_labels = spARI_example_data$c1_labels
c2_labels = spARI_example_data$c2_labels
coords = spARI_example_data$coords
```

Run "spARI" function to compute spRI and spARI.

```{r}
res_value1 = spARI(r_labels=true_labels, c_labels=c1_labels, coords=coords)
res_value2 = spARI(r_labels=true_labels, c_labels=c2_labels, coords=coords)
cat(paste0("1st clustering partition: spRI=", round(res_value1[1], 3), 
           ", spARI=", round(res_value1[2], 3), "\n"))
cat(paste0("2nd clustering partition: spRI=", round(res_value2[1], 3), 
           ", spARI=", round(res_value2[2], 3), "\n"))
```

### Compute spRI and spARI using the sparse distance matrix

In this example, we generate two partitions for 100,000 objects with simulated spatial coordinates to examine the computational efficiency of the `spARI` package for large-scale spatial data.

Since storing a dense matrix of 100,000$\times$100,000 would require more than 16GB of memory, we generate a sparse distance matrix of the objects. 

```{r, message=FALSE, warning=FALSE}
library(FNN)
library(Matrix)

## Define sparse distance matrix generation function
build_symmetric_knn_distance_matrix <- function(coord_mat, k = 5) {
  N <- nrow(coord_mat)
  
  # Find k nearest neighbors for each object (exclude itself)
  knn_res <- FNN::get.knn(coord_mat, k = k)
  i_idx <- rep(1:N, each = k)
  j_idx <- as.vector(t(knn_res$nn.index))
  d_val <- as.vector(t(knn_res$nn.dist))
  
  # Generate sparse matrix (symmetric)
  D_temp <- sparseMatrix(
    i = i_idx,
    j = j_idx,
    x = d_val,
    dims = c(N, N)
  )
  
  D1 <- as(D_temp, "dgTMatrix")
  D2 <- as(t(D_temp), "dgTMatrix")
  
  i_all <- c(D1@i, D2@i)
  j_all <- c(D1@j, D2@j)
  x_all <- c(D1@x, D2@x)
  
  # Combine and aggregate max
  df <- data.frame(i = i_all + 1, j = j_all + 1, x = x_all)  # +1 for R indexing
  df_agg <- aggregate(x ~ i + j, data = df, FUN = max)
  
  # Create symmetric sparse matrix
  D_sym <- sparseMatrix(i = df_agg$j, j = df_agg$i, x = df_agg$x)
  
  return(D_sym)
}

## Generate sparse distance matrix
set.seed(123)
N <- 1e5  # 100,000 objects
coords <- matrix(runif(N*2), N, 2)  # coordinates ranging between 0 and 1
dist_mat <- build_symmetric_knn_distance_matrix(coords, k = 5)

dim(dist_mat)
length(dist_mat@x)
```

**Remark.** For each object, we initially retain only the distances to its $k=5$ nearest neighbors (recorded entries in the distance matrix). However, since the $k$-nearest neighbor relationship is not symmetric (i.e., object $i$ may be among the $k$ nearest neighbors of object $j$, but $j$ is not necessarily among the $k$ nearest neighbors of $i$), the resulting matrix $D_{\text{temp}}$ is asymmetric. To ensure symmetry, we include the distance between two objects if either one is among the $k$ nearest neighbors of the other.

Then we generate the synthetic reference and clustering partitions, with the clustering result containing 10,000 misclustered objects.

```{r}
K = 15
set.seed(123)

# reference partition
true_labels = sample(1:K, N, replace = TRUE)
# clustering partition
c_labels = true_labels
ids = sample(which(true_labels == 1), 5000)
c_labels[ids] = 2
ids = sample(which(true_labels == 3), 5000)
c_labels[ids] = 4
```

Compute spRI and spARI values.

```{r}
library(spARI)

stime = Sys.time()
res = spARI(true_labels, c_labels, dist_mat=dist_mat)
etime = Sys.time()
print(res)
print(etime-stime)
```

### Conducting hypothesis testing

We conduct the permutation test for the observed spARI value of the two clustering partitions in the example data. 

First, we import `spARI` package and load the example data

```{r}
library(spARI)
data("spARI_example_data")
true_labels = spARI_example_data$true_labels
c1_labels = spARI_example_data$c1_labels
c2_labels = spARI_example_data$c2_labels
coords = spARI_example_data$coords
```

Then we carry out the permutation test for the clustering partitions "c1_labels" and "c2_labels". By default, the number of permutations is set to 100.

```{r}
perm_test(r_labels=true_labels, c_labels=c1_labels, coords=coords, 
          use_parallel=FALSE)
perm_test(r_labels=true_labels, c_labels=c2_labels, coords=coords, 
          use_parallel=FALSE)
```

Both the p-values are identical to zero, indicating that the observed spARI value is significantly larger than zero, and the null hypothesis "spARI=0" is rejected.

### 4. Use `SpatialExperiment` object as input

The R package is also compatible with `SpatialExperiment` object, which is a commonly used R object to store ST data. 

First, we generate a synthetic ST data stored in `SpatialExperiment` object.

```{r, message=FALSE, warning=FALSE}
library(SpatialExperiment)
library(S4Vectors)

set.seed(123)
count_matrix <- matrix(
  sample(0:10, 100, replace = TRUE),
  nrow = 10,  # 10 genes
  ncol = 10   # 10 spots
)
rownames(count_matrix) <- paste0("gene", 1:10)
colnames(count_matrix) <- paste0("spot", 1:10)

# Construct gene annotations (rowData)
gene_annotation <- DataFrame(
  gene_id = rownames(count_matrix),
  gene_name = paste0("Gene_", 1:10)
)

# Construct spot metadata (colData)
spot_metadata <- DataFrame(
  spot_id = colnames(count_matrix),
  sample_id = rep("sample1", 10),
  cell_type = c(2, 3, 2, 2, 1, 3, 1, 3, 1, 3),
  cluster = c(2, 3, 2, 3, 1, 3, 3, 3, 1, 3)
)

# Construct spatial coordinates (spatialCoords)
coords_matrix <- cbind(
  x = runif(10, min = 0, max = 100),
  y = runif(10, min = 0, max = 100)
)
rownames(coords_matrix) <- colnames(count_matrix)

# Construct the SpatialExperiment object
spe <- SpatialExperiment(
  assays = list(counts = count_matrix),
  rowData = gene_annotation,
  colData = spot_metadata,
  spatialCoords = coords_matrix
)
```

Then we use the SpatialExperiment object to calculate the spRI and spARI values and conduct the permutation test.

```{r}
library(spARI)
spARI(spe=spe)
perm_test(spe=spe, use_parallel=FALSE)
```
